Learning Action Models from State Observations

### 2018-01-25 doc/ijcai-18/observations-learning.tex
section 4 of ijcai18 paper updated

\section{Evaluation of \strips\ action models}
How well a \strips\ action model $\Xi$ explains a given sequence of observations $\mathcal{O}$ can be assessed according to the amount of {\em edition} required by $\Xi$ to produce $\mathcal{O}$ (in the extreme, if $\Xi$ perfectly explains $\mathcal{O}$, then no {\em edition} of the model is necessary).

@@

We can now formalize an edit distance that quantifies how dissimilar two given \strips\ action models are. This distance is symmetric and satisfies the {\em metric axioms} provided that the two {\em \strips\ edit operations} have the same positive cost.

@@

\begin{definition}
Let $\Xi$ be an action model built from the set of possible elements $F_v$. The \textbf{maximum edit distance} is $\delta(\Xi,*)=\sum_{\xi\in\Xi} 3|F_v(\xi)|$ and sets an upperbound on the edit distance from $\Xi$ to any other \strips\ action model definable within $F_v$.
\end{definition}

@@

\begin{definition}
Given an action model $\Xi$ built from the set of possible elements $F_v$ and a sequence of observations $\mathcal{O}$. The {\bf observation edit distance}, denoted by  $\delta(\Xi,\mathcal{O})$, is the minimal edit distance from $\Xi$ to a model $\Xi'$ such that: (1) $\Xi'$ is also definable within $F_v$ and (2), $\Xi'$ can produce a plan $\pi=\tup{a_1, \ldots, a_n}$ that induces $\mathcal{O}=\tup{s_0, s_1, \ldots, s_n}$, the given observations sequence; i.e., \[\delta(\Xi,\mathcal{O})=\min_{\forall \Xi' \rightarrow \mathcal{O}} \delta(\Xi,\Xi')\]
\end{definition}

Because its semantic nature, the {\em observation edit distance} is robust to learning episodes where actions are reformulated and still compliant with the inputs (e.g. the {\em blocksworld} operator {\small\tt stack} could be {\em learned} with the preconditions and effects of the {\small\tt unstack} operator and vice versa or the roles of the action parameters could be interchanged).

Furthermore the {\em observation edit distance} enables the recognition of \strips\ action models. The idea, taken from {\em plan recognition as planning}~\cite{ramirez2009plan}, is to map distances into likelihoods. We can map the {\em observation edit distance} into a likelihood with the following expression $P(\mathcal{O}|\Xi)=1-\frac{\delta(\Xi,\mathcal{O})}{\delta(\Xi,*)}$. The larger the {\em observation edit distance} the lower this likelihood.

@@

\subsection{Computing the observation edit distance}
Our compilation can be extended to compute the {\em observations edit distance}. In this case the tuple $\Lambda=\tup{\Psi,\Xi,\mathcal{O}}$ is not representing a learning task but the task of editing the \strips\ action model $\Xi$ to produce the observations $\mathcal{O}$. The extended compilation outputs a classical planning task $P_{\Lambda}'=\tup{F_{\Lambda},A_{\Lambda}',I_{\Lambda}',G_{\Lambda}}$ where:
\begin{itemize}
\item $F_{\Lambda}$ and $G_{\Lambda}$ are defined as in the previous compilation.
\item $I_{\Lambda}'$ contains the fluents from $F$ that encode $s_0$ (the first observation) and $mode_{prog}$ set to true. In addition, the given \strips\ action model $\Xi$ is now encoded in the initial state. This means that the fluents $pre_f(\xi)/del_f(\xi)/add_f(\xi)$, with $f\in F_v(\xi)$, hold in the initial state iff they appear in the given $\Xi$ model.
\item $A_{\Lambda}'$, comprises the same three kinds of actions of $A_{\Lambda}$. The Actions for {\em applying} an already programmed operator schema and the actions for {\em validating} an observation {\tt\small $1\leq i\leq n$} are defined exactly as in the previous compilation. The only difference is the actions for {\em programming} operator schema $\xi\in\Xi$ that include now actions for adding a precondition and removing a {\em negative} or {\em positive} effect. They now implement the two {\em edit operations}. 
\end{itemize}

A solution to the classical planning task resulting from the extended compilation is a sequence of actions that edits the \strips\ action model $\Xi$ and validates the edited model on the given observations $\mathcal{O}$. Since we are computing the {\em observations edit distance}, we are not interested in the edited model but in the number of {\em edit operations} required to produce a model validable in the given observations. The {\em observation edit distance} is exact if $P_{\Lambda}'$ is optimally solved (according to the number of edit actions), and is approximate if $P_{\Lambda}'$ is solved with a satisfying planner or furthermore, if what is solved is not $P_{\Lambda}'$ but its relaxation, e.g. the {\em delete relaxation}~\cite{bonet2001planning}.

>>>>

\section{Evaluation of \strips\ action models}
We assess how well a \strips\ action model $\Xi$ explains a given sequence of observations $\mathcal{O}$ according to the amount of {\em edition} that is required by $\Xi$ to produce $\mathcal{O}$. In the extreme, if $\Xi$ perfectly explains $\mathcal{O}$, no model {\em edition} is necessary.

@@

We can now formalize an edit distance that quantifies how dissimilar two given \strips\ action models are. The distance is symmetric and meets the {\em metric axioms} provided that the two {\em \strips\ edit operations} have the same positive cost.

@@

\begin{definition}
The \textbf{maximum edit distance} of an \strips\ action model $\Xi$ built from the set of possible elements $F_v$ is $\delta(\Xi,*)=\sum_{\xi\in\Xi} 3|F_v(\xi)|$ and sets an upperbound on the distance from $\Xi$ to any other \strips\ action model definable within $F_v$.
\end{definition}

@@

\begin{definition}
Given an action model $\Xi$ built from the set of possible elements $F_v$ and a sequence of observations $\mathcal{O}$. The {\bf observation edit distance}, denoted by  $\delta(\Xi,\mathcal{O})$, is the minimal edit distance from $\Xi$ to any model $\Xi'$ such that: (1) $\Xi'$ is also definable within $F_v$ and (2), $\Xi'$ can produce a plan $\pi=\tup{a_1, \ldots, a_n}$ that induces $\mathcal{O}=\tup{s_0, s_1, \ldots, s_n}$, the given observations sequence; i.e., \[\delta(\Xi,\mathcal{O})=\min_{\forall \Xi' \rightarrow \mathcal{O}} \delta(\Xi,\Xi')\]
\end{definition}

Because its semantic nature, the {\em observation edit distance} is robust to learning episodes where actions are {\em reformulated} and still compliant with the learning inputs. For instance a learning episode can interchange the roles of operators with matching headers, or the roles of action parameters with matching types (e.g. the {\em blocksworld} operator {\small\tt stack} can be {\em learned} with the preconditions and effects of the {\small\tt unstack} operator and vice versa).

Furthermore, the {\em observation edit distance} enables the recognition of \strips\ action models. The idea, taken from {\em plan recognition as planning}~\cite{ramirez2009plan}, is to map distances into likelihoods. We can map the {\em observation edit distance} into a likelihood with the following expression $P(\mathcal{O}|\Xi)=1-\frac{\delta(\Xi,\mathcal{O})}{\delta(\Xi,*)}$. The larger the {\em observation edit distance} the lower this likelihood.

@@

\subsection{Computing the observation edit distance}
Our compilation is extensible to compute the {\em observations edit distance}. A solution to the planning task resulting from the extended compilation is a sequence of actions that:
\begin{enumerate}
\item {\bf Edits the \strips\ action model $\Xi$}. A solution plan starts with a {\em prefix} that edits $\Xi$.
\item {\bf Validates the edited model $\Xi$ in $\mathcal{O}$}. The solution plan continues with a postfix that validates the edited model on the given observations $\mathcal{O}$.
\end{enumerate}

In this case the tuple $\Lambda=\tup{\Xi,\Psi,\mathcal{O}}$ does not represent a learning task but the task of editing the \strips\ action model $\Xi$ to produce the observations $\mathcal{O}$. The extended compilation outputs a classical planning task $P_{\Lambda}'=\tup{F_{\Lambda},A_{\Lambda}',I_{\Lambda}',G_{\Lambda}}$:
\begin{itemize}
\item $F_{\Lambda}$ and $G_{\Lambda}$ are defined as in the previous compilation.
\item $I_{\Lambda}'$ contains the fluents from $F$ that encode $s_0$ (the first observation) and $mode_{prog}$ set to true. In addition, the given \strips\ action model $\Xi$ is now encoded in the initial state. This means that the fluents $pre_f(\xi)/del_f(\xi)/add_f(\xi)$, with $f\in F_v(\xi)$, hold in the initial state iff they appear in the given $\Xi$ model.
\item $A_{\Lambda}'$, comprises the same three kinds of actions of $A_{\Lambda}$. The actions for {\em applying} an already programmed operator scheme and the actions for {\em validating} an observation are defined exactly as in the previous compilation. The only difference here is the actions for {\em programming} operator schemes $\xi\in\Xi$ that now implement the two {\em edit operations} (i.e. include actions for adding a precondition and for removing a negative/positive effect).  
\end{itemize}

Since we are computing the {\em observations edit distance}, we are not interested in the resulting edited model but in the number of {\em edit operations} required to produce a model validable in the given observations. The {\em observation edit distance} is exactly computed if $P_{\Lambda}'$ is optimally solved (according to the number of edit actions), and is approximate if $P_{\Lambda}'$ is solved with a satisfying planner or furthermore, if what is solved is not $P_{\Lambda}'$ but its relaxation, e.g. the {\em delete relaxation}~\cite{bonet2001planning}.
